\documentclass[12pt,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=0.76in,columnsep=0.25in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}
\setlength{\emergencystretch}{2em}
\hfuzz=2pt
\title{\vspace{-1.2em}From Robust Skill Selection to Recurrent Robot-Soccer Team Play}
\author{Research software report}
\date{23 July 2026}

\begin{document}
\maketitle

\begin{abstract}
We study zero-shot transfer of decentralized, parameter-shared robot-soccer policies from a fast
abstract simulator to an independent Pymunk rigid-body implementation. Phase 2 compared
equal-budget nominal, uniformly randomized, and failure-directed MAPPO over three independent
training seeds. Failure-directed randomization reliably improved the 18-profile Pymunk mean
($0.515\pm0.045$ versus $0.307\pm0.075$ nominal) and the delay--localization grid
($0.455\pm0.077$ versus $0.315\pm0.121$). It nevertheless failed the locked gate: abstract
intercept competence fell from $0.763\pm0.065$ to $0.427\pm0.068$, and cooperative success was
zero in all three failure-directed actors. Phase 2 therefore establishes a real robustness
trade-off, not a successful team-play policy. Phase 3 is a targeted redesign: calibrated 2-v-2
and 3-v-2 tasks, a recurrent shared actor and recurrent centralized critic, staged nominal
training, and competence-constrained failure-directed fine-tuning. The failed Phase 2 actor is
not promoted. Phase 3 development seed 3 passed the full calibration gate and completed all four
nominal stages. Its frozen Stage-D policy passed four of five nominal-transfer checks, including
$0.61$ pass-required cooperative success and $0.75$ 3-v-2 success, but failed locked Gate B
because 2-v-2 open success was $0.49<0.55$. CC-FDR is therefore correctly blocked.
\end{abstract}

\section{Research contract}
The project asks whether a compact cooperative policy can acquire behavior that is simultaneously
useful, transferable, and interpretable. The concrete claim under test is:
\begin{quote}
Can recurrent, parameter-shared MAPPO learn genuine multi-attacker play in an abstract simulator
and preserve that nominal competence while failure-directed training improves frozen-policy
transfer to independent rigid-body physics?
\end{quote}
The intended contribution is a controlled mechanism result rather than a leaderboard. We use
macro-actions above an assumed locomotion layer, two independently implemented dynamics backends,
fixed evaluation seeds, and auditable artifacts. Pymunk transfer is a sim-to-sim proxy, not a
physical-robot claim. Full 11-v-11 play, pixel control, and self-play populations are deliberately
outside the present scope.

The RoboCup-facing north star is a learned high-level football policy integrated above
perception/localization and walking, turning, dribbling, and kicking primitives:
\[
\begin{gathered}
\text{high-level football policy}\\[-0.2em]
\downarrow\\[-0.2em]
\text{perception/localization}\\[-0.2em]
\downarrow\\[-0.2em]
\text{motion primitives}\\[-0.2em]
\downarrow\\[-0.2em]
\text{robot platform}
\end{gathered}
\]
This deliberate hierarchy tests centralized-training/decentralized-execution decision making,
partial observability, recurrence, cooperation, and adversarial variation; it does not train
humanoid joints or imply that success here alone wins RoboCup.

This scope follows the staged approach used by robot-football systems: isolate useful drills and
high-level coordination before adding full-match complexity \cite{liu2022teamplay,kurach2020football,
su2025quadruped}. PPO supplies the clipped on-policy objective \cite{schulman2017ppo}, MAPPO supplies
centralized training with decentralized execution \cite{yu2022mappo}, and domain randomization
supplies a transfer intervention \cite{tobin2017domain,mehta2020active}. The experimental standard
is independent training seeds and seed-level uncertainty, not confidence intervals over repeated
episodes from one actor \cite{agarwal2021precipice}.
Recent UW--Madison work independently supports this boundary: abstract multi-robot policies can
transfer through separately obtained perception and motion modules \cite{labiosa2025abstract},
while learned soccer decisions can be integrated within a classical robotics stack using multiple
fidelity levels \cite{labiosa2025stack}. We adopt those motivations, not their exact simulator,
robot, or deployment result.

\section{Phase 1: engineering gate}
Phase 1 corrected playable-boundary behavior, possession normalization, profile aggregation, and
baseline semantics. Its final 100-episode cells produced abstract/Pymunk success of
$0.03/0.49$ for random, $0.15/0.68$ for double-chase, and $0.28/0.66$ for role-based play.
Role-based control greatly reduced redundant chasing, establishing policy sensitivity. The high
Pymunk random success also exposed a limitation: that intercept setup was permissive and could not
serve as a meaningful team-play calibration task. Phase 3 therefore introduces a separate,
stricter calibration gate rather than reusing those numbers.

\section{Phase 2 protocol}
Phase 2 trained nominal, uniform-DR, and failure-directed MAPPO for the same budget. Each actor was
frozen and evaluated on untouched confirmatory seed ranges: abstract intercept, nominal Pymunk
transfer, an 18-profile Pymunk suite, a delay--localization grid, and a pass-needed cooperation
probe. Failure-directed sampling increased probability for observed failures while retaining a
nominal rehearsal fraction.

The executed notebook contains all nine complete runs and the suite-safe comparison output.
The lightweight Mac archive currently contains only the three seed-0 raw run directories; the
seed-1 and seed-2 values below are therefore transcribed into a locked JSON summary with explicit
notebook/cell provenance. This distinction matters: the numerical evidence is recoverable and
auditable, but the local archive cannot independently regenerate every CSV until those six run
directories are restored from Drive.

\IfFileExists{reports/generated_results.tex}
  {\input{reports/generated_results.tex}}
  {\input{generated_results.tex}}

\section{What Phase 2 actually established}
\subsection{Robustness redirected successfully}
Failure-directed MAPPO beats both controls on the Pymunk profile mean in every matched training
seed. Its aggregate profile mean is $0.515$, compared with $0.353$ for uniform DR and $0.307$ for
nominal MAPPO. Its grid AUC is also largest: $0.455$ versus $0.327$ and $0.315$. This corrects the
single-seed pilot story. The earlier seed-0 delay collapse did not replicate across training
seeds, so a delay-only ablation is not authorized. The defensible mechanism claim is that
failure-directed sampling consistently reallocates capacity toward the evaluated perturbation
distribution.

\subsection{The overall scientific gate failed}
The same actors lose too much nominal competence. Mean abstract intercept success falls by
$0.337$, far beyond the allowed $0.10$ non-inferiority margin. This is not small statistical
noise: all three FDR seeds lie between $0.35$ and $0.48$, while all nominal seeds lie between
$0.70$ and $0.83$. Uniform DR is unstable, including one strong abstract seed paired with weak
profile/grid performance. These patterns indicate objective interference and training-seed
sensitivity, not a universally robust policy.

The cooperation result is more decisive. Failure-directed cooperative success is $0$ for all
three seeds. Uniform DR has one $0.32$ seed and two zeros, which is neither stable nor evidence of
a repeatable strategy. A two-attacker intercept task permits direct approach-and-shoot behavior;
passing is therefore optional and cannot be inferred from goal rate or spatial separation.

\subsection{Critical decision}
Phase 2 is complete as a diagnosis and failed as a graduation gate. Repeating the same feed-forward
jobs is unlikely to fix structural partial observability or the competence trade-off. Moving
directly to harder physics would confound learning failure with simulator difficulty. The next
phase must first make cooperation necessary, supply memory, and constrain robust fine-tuning.

\section{Phase 3 design}
\subsection{Calibrated team-play tasks}
Phase 3 uses a fixed maximum interface of three attackers and three defender slots, with padded
egocentric observations and explicit presence/action masks. Active teams are 2-v-2 or 3-v-2.
Scenarios include open play, pass-required 2-v-2, open 3-v-2, and high-pressure 3-v-2. Deterministic
entity ordering and a backward-compatible 2-v-1 adapter prevent observation identity from changing
across team sizes.

Before learning, scripted random, double-chase, and role policies are evaluated in abstract and
Pymunk. The locked calibration requires, among other checks, Pymunk random success at most $0.05$,
double-chase at most $0.20$, role success at least $0.15$ above random and between $0.20$ and
$0.55$, direct-shot double-chase success in pass-required play at most $0.05$, median successful
time at least eight seconds, and no cell above $0.85$. Failure blocks expensive training; thresholds
are not weakened to manufacture a pass.

\subsection{Recurrent centralized training}
The shared actor uses a $[192,192]$ encoder and a 256-unit GRU. Each environment/attacker slot owns
its recurrent state; state resets on an actual episode boundary, not a match restart. A recurrent
central critic consumes the padded global state. Rollouts retain initial hidden states, validity
masks, termination masks, and contiguous time chunks. PPO minibatches use truncated
back-propagation through time with configurable sequence length and burn-in; masked logits make
illegal macro-actions impossible in training, evaluation, baselines, and video.

The recurrent design addresses two Phase 2 weaknesses directly. Delayed actions and intermittent
messages make a frame stack an imperfect state estimator, while pass-and-receive behavior requires
temporal credit across possession changes. Memory does not guarantee cooperation, so the task and
metrics remain the primary safeguards.

\subsection{Staged nominal curriculum}
Nominal training proceeds through four resumable stages:
\begin{enumerate}
\item 2-v-2 open play, target budget $0.5$ million steps;
\item 2-v-2 pass-required play, cumulative $1.0$ million;
\item mixed 2-v-2/3-v-2 open play, cumulative $1.5$ million; and
\item mixed open/press team play, cumulative $2.0$ million.
\end{enumerate}
Checkpoints carry model, optimizer, normalizer, recurrent curriculum, RNG, and stage state. A smoke
configuration exercises every pathway without pretending to be a trained result.

\subsection{Competence-constrained failure direction}
Failure-directed fine-tuning starts only from the best nominal recurrent checkpoint. Let $p_0$ be
the nominal rehearsal probability, $R_N$ the current nominal score, $R_N^\star$ its frozen
reference, and $\delta=0.10$ the allowed loss. The controller enforces $p_0\geq0.35$. If
$R_N<R_N^\star-\delta$, it increases rehearsal and reduces failure focusing. Checkpoint selection
tracks nominal, cooperation, and configurable composite objectives separately. This makes the
nominal constraint operational rather than a post-hoc apology.

Sampling is explicitly partitioned into nominal-football rehearsal, pass-required cooperation,
and randomized robustness groups; raising nominal rehearsal cannot silently remove the dedicated
cooperation stream.

Reward shaping provides controlled-reception, pass-advance, and pass-to-goal credit but no reward
for merely selecting ``pass.'' Cooldowns and possession-transition checks prevent stationary pass
credit from repeating without a new transition; executed results must still be audited for
cyclic pass churning. Evaluation reports the unshaped outcomes: goals, valid attempts, completed
receptions, possession chains, expected-threat gain, redundant chasing, and worst-profile success.

\section{Decision gates and run order}
Gate A is calibration. Gate B is nominal team play: the recurrent actor must beat role-based
control and meet the locked Pymunk thresholds: 2-v-2 open success $0.55$, 3-v-2 open success
$0.35$, pass-required cooperation $0.30$, median successful sequence duration eight seconds, and
at least two meaningful actions. Gate C is constrained robustness:
failure-directed fine-tuning must improve the declared robustness aggregate while staying within
the nominal floor. Development uses seed 3. Only after mechanisms and thresholds are frozen are
seeds 4, 5, and 6 used for final estimates.

The sequence is therefore: run unit/equivalence tests; run the calibration smoke and then the full
calibration gate; run throughput diagnostics; train staged nominal seed 3; evaluate both physics
backends and inspect videos; fine-tune CC-FDR from the selected nominal checkpoint; then freeze the
protocol and run seeds 4--6. Harder simulation or hardware is considered only if all three gates
pass.

\section{Phase 3 development result}
The full 100-episode-per-cell calibration passed every locked check. In Pymunk 2-v-2 open play,
random, double-chase, and role-based success were $0.01$, $0.03$, and $0.26$; double-chase scored
zero in pass-required play. The benchmark is therefore difficult, policy-sensitive, and protected
against the direct-shot control. Throughput reached $337.45$ transitions/s at 128 environments and
then plateaued, so 128 was a reasonable operating point rather than a claimed algorithmic result.

Seed-3 nominal training completed $2{,}007{,}040$ cumulative steps. Abstract validation progressed
to $1.00$ nominal and cooperative success, with finite optimization, final entropy $0.517$,
approximate KL $0.0011$, and explained variance $0.857$. External Pymunk probes were less
saturated: Stage A achieved $0.43$ in 2-v-2 open and Stage B achieved $0.37$ in pass-required
play. The frozen Stage-D Gate-B results were:
\begin{center}
\scriptsize
\begin{tabular}{lrrr}
\toprule Check & Observed & Required & Pass \\
\midrule
2-v-2 open success & 0.49 & 0.55 & no \\
3-v-2 open success & 0.75 & 0.35 & yes \\
pass-required cooperation & 0.61 & 0.30 & yes \\
successful duration (s) & 17.2 & 8.0 & yes \\
meaningful actions & 6.19 & 2.0 & yes \\
\bottomrule
\end{tabular}
\normalsize
\end{center}
This is a near miss in aggregate competence and a substantial improvement in identifiable
cooperation, but Gate B is conjunctive. It did not pass.

The aggregate hides a serious style gap. In 2-v-2 open play, success was $1.00$ against press,
$0.483$ against zonal, $0.333$ against lane-block, and $0.053$ against predictive defense.
Lane-block episodes averaged 93.1 pass attempts and a high shaped return despite only $0.333$
success.

\subsection{Post-sync forensic audit}
The raw episode rows make the lane-block failure more specific. The 18 failed lane-block matches
averaged 106.5 pass attempts, 105.8 receptions, 0.56 shots, zero shots on target, and return
$35.82$. The nine successes averaged 66.4 passes, 64.2 receptions, 5.33 shots, and return $37.34$.
Controlled-reception reward alone contributed about $37.0$ in the typical failed match. Thus a
long, non-progressing circulation can nearly match the return from scoring; this is objective
misalignment, not merely an implementation-crash hypothesis. Predictive-defense failures are
different: they average 11.61 shots, zero on target, 7.39 out-of-bounds events, and return
$-19.18$. Zonal failures show a third pattern, with 62.1 collision events and little progression.
The counter increments once per macro-action decision with attacker overlap, not only at contact
onset, so sustained contact can contribute repeatedly. Stage R therefore leaves the collision
coefficient unchanged. An architecture sweep would conflate three behavioral failures that the
current network already has enough capacity to express.

Six synchronized 20-second clips visibly show carrier/support role separation, completed passes,
and scoring/restarts. This is nontrivial team play, not a trivial direct-shot baseline. The clips
are nevertheless partial: they stop before the 50-second match terminal, contain no final metrics,
and the recorder overwrites its manifest on each invocation. They support a behavior claim, but
cannot resolve the failed gate or quantify style robustness.

\subsection{Immediate experiment: nominal mastery Stage R}
The implemented, scientifically unrun intervention is one explicit repair stage, not more
unchanged Stage-D training. This
choice is consistent with evidence that MAPPO can be strong without elaborate architecture
changes \cite{yu2022mappo}, that curricula should concentrate on levels with current learning
value \cite{jiang2021plr}, and that shaped rewards can be optimized at the expense of the intended
outcome \cite{forbes2025shaping}. The evaluator and recorder now support forced defender style,
merged/deduplicated manifests, terminal/full-match capture, seed-specific replay, and current
metrics for partial clips. The frozen Stage-D actor is assigned a declared $4$-style
$\times$ $2$-scenario
$\times$ $2$-backend audit with 100 episodes per cell.

Stage R warm-starts the Stage-D actor, critic, and normalizers while resetting optimizer,
scheduler, rollout, and Stage-R counters. Its one authorized scientific run is 500,000
abstract-simulator steps, initially at $10^{-4}$ learning rate with a
$3\times10^{-5}$ floor. Scenario probabilities will be 0.50 2-v-2 open, 0.20 pass-required,
0.15 3-v-2 open, and 0.15 3-v-2 press; defender probabilities will be 0.35 lane-block,
0.35 predictive, 0.20 zonal, and 0.10 press. Controlled-reception reward is exactly zero from the
first Stage-R transition. A possession-chain high-water mark pays pass advance only for a newly
reached forward frontier; backward and A--B--A--B revisits pay zero, while the next new frontier
pays only its incremental excess. Goal, pass-to-goal, and ordinary potential shaping remain.
Deterministic machine-readable invariants lock these semantics. Checkpoint selection uses
held-out, style-stratified abstract football outcomes rather than shaped return or Pymunk feedback.

The repair evaluation uses a new seed base and retains every original Gate-B threshold. It also
predeclares three safeguards: pass-required success at least $0.55$, 3-v-2 success at least
$0.70$, and 2-v-2 open worst-style success at least $0.25$ with the lane/predictive mean at least
$0.40$. For styles containing both outcomes, successful return must exceed failed return by at
least five. These are Stage-R development safeguards, not a retrospective rewrite of Gate B.
Only a pass authorizes the renumbered CC-FDR Section 39. If Stage R passes in abstract but fails Pymunk, the
next intervention is a calibrated abstract dynamics envelope; if it fails both, rerun A--D plus R
from scratch with the corrected reward schedule. Procedural task variation is retained because
fixed, easily memorized tasks can overstate closed-loop competence \cite{ellis2023smacv2}.

\section{Experiment ledger and evidence boundary}
Table~\ref{tab:ledger} separates completed evidence from executable but unrun work. This is
especially important in reinforcement learning, where a functioning training loop, an attractive
video, and a statistically supported policy claim are three different accomplishments.
\begin{table*}[t]
\centering
\small
\begin{tabular}{p{0.18\textwidth}p{0.20\textwidth}p{0.34\textwidth}p{0.18\textwidth}}
\toprule Phase & Status & Durable evidence & Permissible conclusion \\
\midrule
Phase 1 & complete & corrected simulator tests; 600 held-out baseline episodes; readiness JSON &
the original 2-v-1 interface is runnable and policy-sensitive, but permissive in Pymunk \\
Phase 2 pilot & complete & three equal-budget seed-0 actors; suite-specific CSV/JSON; videos &
failure direction is promising but a single actor is exploratory \\
Phase 2 confirmation & complete & nine actors; locked per-seed summary; untouched evaluation
ranges; automated gate & profile/grid robustness improves, nominal competence and cooperation fail \\
Phase 3 software & implemented & recurrent smoke/resume/export tests; full calibration; staged
training and fail-closed gate tooling & the protocol is executable and retained failed-gate
evidence \\
Phase 3 Stage R & implemented, unrun & reward invariants; fixed-style R0/Gate-B-R tooling;
warm-start and video tests; preserved Stage A--D evidence & cooperation is real, but no Stage-R
performance claim exists before the Colab run \\
Phase 3 confirmation & disabled & future new seeds 4, 5, 6 after the protocol freezes & no
multi-seed recurrent claim \\
\bottomrule
\end{tabular}
\caption{Experiment ledger. ``Implemented'' is never substituted for ``empirically passed.''}
\label{tab:ledger}
\end{table*}

\subsection{Required Phase 3 artifacts}
Each training job must save the resolved configuration, source commit, environment and recurrent
schema versions, optimizer and normalization state, stage/global counters, RNG states, scalar CSV,
detailed log, validation JSON, render check, and final checkpoint. CC-FDR additionally saves its
nominal reference, nominal floor, rehearsal trajectory, profile probabilities, constraint margins,
and guard events. Evaluation jobs write episode rows before aggregating. A table is reportable only
when those rows can reproduce it.

Videos use matched, declared seeds and carry scenario, simulator, defender style, actions,
possession, profile, clock, and score overlays. They are behavior audits: a positive clip cannot
replace the full seed set, and a failure clip is retained rather than searched away. The expected
Phase 3 set includes 2-v-2 pass-required behavior, 3-v-2 pressure, a randomized robustness case,
and a representative failure.

\subsection{Why the redesign is minimal}
The intervention changes three coupled items and holds the rest stable. First, recurrence addresses
delayed/hidden state. Second, pass-required geometry makes the missing behavior identifiable.
Third, the competence constraint targets the demonstrated FDR trade-off. We do not add a second RL
algorithm, learned-opponent population, pixel encoder, or joint-level controller. Those additions
would multiply explanations before the basic cooperation/transfer mechanism is known.

The comparison remains paired: recurrent nominal and CC-FDR share architecture, nominal
pretraining, macro-actions, environment semantics, stage budget accounting, evaluation seeds, and
checkpoint rules. CC-FDR alone changes the fine-tuning distribution subject to its rehearsal
guard. This makes a future difference interpretable as a curriculum effect rather than an
architecture or compute advantage.

\subsection{Failure-contingent next actions}
If Gate A fails, modify task geometry or scripted behavior and rerun calibration; never alter the
threshold after seeing a learned policy. Gate A passed. Gate B failed with healthy optimization,
and the completed forensic analysis found reward misalignment plus distinct style-specific
failures. The predeclared Stage-R reward, curriculum, diagnostics, and safeguards are now
implemented but unrun; do not change the network or tune PPO simultaneously. Run R0, save the
invariant artifact, train exactly one Stage-R job, run abstract validation, and then Gate B-R.
If abstract competence does not transfer, add a calibrated abstract dynamics envelope before
robust fine-tuning. If Gate C later improves robustness but violates the floor, treat the
competence guard as insufficient; increase rehearsal or stop fine-tuning, then rerun seed 3 before
any final seed. If all gates pass, freeze code/config hashes before seeds 4--6.

\section{Threats to validity}
Macro-actions omit low-level contact control. Scripted defenders may expose exploitable regularity.
Pymunk and abstract physics share task geometry even though their integration differs. Adaptive
sampling can chase noisy failures, while recurrent PPO introduces optimization sensitivity.
Pass-required geometry can itself overfit a scripted solution. We mitigate these threats with
mixed defender styles, held-out combinations, direct-shot controls, matched seeds, exact artifact
provenance, videos, and separate nominal/robustness checkpoints. Three final seeds remain a modest
sample; conclusions must report seed-level dispersion.

\section{Present status}
The project is advancing toward the stated objective rather than merely accumulating benchmarks:
the calibrated task is nontrivial, recurrent nominal training is stable, and the policy now
exhibits substantial pass-required cooperation in independent Pymunk physics. However, Phase 3
development is not complete. Gate B failed one locked competence threshold and exposed severe
defender-style variation plus a confirmed mismatch between controlled-reception reward and
scoring. Nominal-mastery Stage R is implemented and smoke-tested but scientifically unrun. The
current decision is \emph{run the one predeclared Stage-R protocol}; unchanged
Stage-D extension, CC-FDR, harder simulation, architecture search, and seed-4/5/6 confirmation
remain disabled. This preserves the end objective---robust transfer without loss of useful
cooperative football---instead of optimizing a favorable subset of metrics.

\bibliographystyle{plain}
\IfFileExists{reports/references.bib}
  {\bibliography{reports/references}}
  {\bibliography{references}}
\end{document}
