\documentclass[12pt,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[margin=0.80in,columnsep=0.25in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}
\setlength{\emergencystretch}{2em}
\title{\vspace{-1.2em}Robot-Soccer Transfer: Technical Surrogate and Runbook}
\author{Auditable implementation notes}
\date{23 July 2026}

\begin{document}
\maketitle

\section{Purpose and truth status}
This document is the technical companion to the public report. It records exact evidence,
limitations, interfaces, gates, and the intended execution order. It deliberately distinguishes
implemented machinery, smoke validation, and completed scientific experiments.

\textbf{Completed evidence:} Phase 1 baseline gate, Phase 2 three-seed confirmation, and Phase 3
seed-3 calibration plus nominal Stages A--D.
\textbf{Current decision:} Phase 2 failed overall because nominal competence and cooperation gates
failed. Phase 3 Gate A passed, but Gate B failed one of five locked checks: Pymunk 2-v-2 open
success was $0.49<0.55$. Stage R is implemented and smoke-tested but scientifically unrun.
\textbf{Not authorized:} CC-FDR, Gate C, harder physics, or final seed-4/5/6 evaluation.

The deployment-facing abstraction is intentional:
\[
\begin{gathered}
\text{learned football policy}\\[-0.2em]
\downarrow\ \text{perception/localization}\\[-0.2em]
\downarrow\ \text{motion primitives}\\[-0.2em]
\downarrow\ \text{robot platform}
\end{gathered}
\]
This repository studies the first layer, not NAO/humanoid joint control. The separation follows
recent UW--Madison evidence on abstract multi-robot policy transfer and RL embedded in a
multi-fidelity classical soccer stack, as cited in the main report; it does not claim an identical
implementation or physical result.

\section{Phase 2 provenance}
The authoritative Phase 2 aggregate is printed in cell 22 of
\path{notebooks/phase2_training_transfer_report.ipynb}. Nine runs completed:
\begin{itemize}
\item seed 0: \texttt{002437 nominal}, \texttt{020456 uniform}, \texttt{033215 FDR};
\item seed 1: \texttt{082453 nominal}, \texttt{095311 uniform}, \texttt{112147 FDR};
\item seed 2: \texttt{151725 nominal}, \texttt{175954 uniform}, \texttt{192845 FDR}.
\end{itemize}
The local lightweight archive currently contains only the seed-0 directories. The locked file
\path{reports/phase2_confirmatory_summary.json} records the exact per-seed notebook values,
missing local paths, seed bases, baseline source, and decision. The LaTeX table is generated from
that JSON by \texttt{python -m scripts.export\_phase2\_report}; it is not a hand-maintained copy.

\IfFileExists{reports/generated_results.tex}
  {\input{reports/generated_results.tex}}
  {\input{generated_results.tex}}

\section{Interpretation of the failed gate}
FDR passes three robustness checks: profile mean superiority, matched-seed direction stability,
and delay--noise grid non-inferiority. It fails abstract non-inferiority by $0.237$ beyond the
allowed margin: nominal $0.7633$, FDR $0.4267$, absolute loss $0.3367$, allowed loss $0.10$.
FDR cooperative success is exactly zero at every seed. The seed-0 delay failure is not stable
across seeds; \texttt{delay\_failure\_replicated=false} and
\texttt{delay\_ablation\_authorized=false}. Do not run or report a delay-specific retraining
ablation under the Phase 2 protocol.

Uniform DR is not a reliable compromise. Its profile standard deviation is $0.210$ and grid
standard deviation is $0.143$. Its only cooperative result, $0.32$ at seed 2, coincides with poor
profile/grid performance. The correct story is objective interference plus training-seed
sensitivity.

\section{Phase 3 fixed interface}
The maximum roster is three attacker slots and three defender slots. Actual active attackers are
two or three; actual active defenders are two. Attacker identifiers are stable and zero-padded.
Every local observation contains:
\begin{itemize}
\item self pose, velocity, heading, possession, and slot identity;
\item ball pose/velocity and recent possession transition;
\item attacking goal and field-relative geometry;
\item two teammate slots with presence bits and a message carrying age and validity;
\item three defender slots with presence bits and relative kinematics;
\item previous/executed action, scenario marker, clock, and action mask.
\end{itemize}
All continuous values use fixed normalizers. Deterministic entity ordering is by stable slot, never
nearest-neighbor rank. Inactive slots are zero with a presence bit of zero. The global critic state
uses fixed world-frame slots and includes scenario/roster masks. The old 2-v-1 interface remains
available through its existing environment; Phase 3 does not silently alter Phase 2 artifacts.

\section{Scenarios and opponent styles}
\scriptsize
\begin{tabular}{p{0.29\columnwidth}p{0.57\columnwidth}}
\toprule Scenario & Intent \\
\midrule
\texttt{2v2 open} & Calibrated open attack with goalie and field defender.\\
\texttt{2v2 pass-required} & Direct lane blocked; a controlled teammate reception is
required before a goal counts as cooperative success.\\
\texttt{3v2 open} & Adds a second off-ball option and padded third attacker.\\
\texttt{3v2 press} & Starts under pressure and mixes defender styles.\\
\bottomrule
\end{tabular}
\normalsize

Defender styles are goalie, lane-block, press, predictive, zonal, and mixed. The goalie stays near
the goal line and tracks the ball; the field defender uses the selected style. ``Mixed'' samples
at reset from an explicit probability map and logs the realized style. Opponent actions are
scripted and excluded from learning-agent tensors.

\section{Calibration gate}
Calibration runs random, double-chase, and role-based policies in abstract and Pymunk open and
pass-required scenarios. Optional 3-v-2 cells are diagnostic. The non-negotiable defaults are:
\begin{align*}
S_{\text{random,Pymunk}}&\leq0.05,\\
S_{\text{double,Pymunk}}&\leq0.20,\\
0.20\leq S_{\text{role,Pymunk}}&\leq0.55,\\
S_{\text{role,Pymunk}}-S_{\text{random,Pymunk}}&\geq0.15,\\
S_{\text{direct,pass-required}}&\leq0.05.
\end{align*}
Successful median duration must be at least eight seconds and no method/scenario/simulator cell may
exceed $0.85$. Missing cells, insufficient episode counts, non-finite metrics, or a permissive
direct shot all fail closed. A 20-episode smoke checks code only. Expensive learning is blocked
until the configured full-episode calibration passes; smoke noise never authorizes training.

\section{Reward and metric contract}
Goal and failure rewards remain team-level. Potential shaping uses ball progress and field control.
Cooperation credit requires events, not action labels:
\begin{enumerate}
\item carrier releases a valid pass after controlled possession;
\item a different attacker controls the ball inside a reception window;
\item the reception advances field position or improves expected threat;
\item a subsequent goal inside the chain receives pass-to-goal credit.
\end{enumerate}
Selecting pass alone yields no cooperation reward. A receiver identity change, minimum ball travel,
cooldown, and one-credit-per-transition rule prevent farming. Logs retain attempted/valid/completed
passes, receptions, chain length, scorer, assister, expected-threat gain, possession switches,
redundant chasing, action entropy, invalid-mask requests, and terminal reason.

Reward schema 1 remains the historical A--D contract. Stage R alone uses schema 2:
controlled-reception events remain metrics but have coefficient zero from transition one.
\texttt{chain\_progress\_high\_water} persists across attacker-to-attacker possession changes.
Only \(\max(0,p_t-\max_{\tau<t}p_\tau)\) receives pass-advance credit; backward receptions and
revisits receive zero. Kickoff, restart, goal, out-of-bounds, defender clear, turnover, and full
reset begin a new chain. Ordinary potential shaping and the single qualifying pass-to-goal event
remain unchanged.

\section{Recurrent MAPPO contract}
The actor is shared across attacker slots:
\[
\begin{aligned}
e_t &= \mathrm{MLP}_{192,192}(o_t),\\
h_t &= \mathrm{GRU}_{256}(e_t,h_{t-1}),\\
\pi(a_t\mid o_{\leq t}) &= \mathrm{softmax}(W h_t).
\end{aligned}
\]
The central value network uses the fixed global state and its own recurrent state. Hidden state has
shape environment $\times$ maximum-attacker-slot $\times256$ for the actor and environment
$\times256$ for the critic. An inactive attacker's hidden state is zero. Environment termination
or truncation resets state; an in-episode match restart does not.

Rollout records observations, global states, sampled and executed actions, action masks, log
probabilities, rewards, values, valid-slot masks, boundaries, and initial hidden states. Minibatches
are contiguous $(\text{environment},\text{slot},\text{time})$ chunks. Configurable burn-in rebuilds
hidden state without loss, then TBPTT applies loss only to valid post-burn-in steps. Advantages are
normalized across valid samples. The implementation warns on high approximate KL, near-total
clipping, non-finite gradients, or actor/critic learning rates below their configured minima.

\section{Stage state and resume}
Nominal stages A--D have cumulative target budgets $0.5$, $1.0$, $1.5$, and $2.0$ million
environment steps. The smoke config uses tiny overrides. Each checkpoint contains stage name,
stage/global counters, actor/critic and optimizer weights, observation/state normalizers,
curriculum state, recurrent-state dimensions, and RNG states. A resume must match observation,
action, roster, recurrent, and scenario schema versions. A stage may warm-start only from its
declared predecessor.

Stage R is a new warm-start experiment, not an A--D resume. It accepts only a Stage-D checkpoint,
copies actor, critic, observation normalizer, and global-state normalizer, and resets optimizers,
scheduler/update/rollout state, and Stage-R counters. Its deterministic training-episode stream
starts at 1000000000. An interrupted Stage-R run can resume only from a Stage-R checkpoint and
restores that seed counter; CC-FDR can warm-start only from a selected Stage-R checkpoint.

\section{Competence-constrained FDR}
CC-FDR is a fine-tuning phase, never a fresh initialization. It stores the frozen nominal reference
$R_N^\star$ and uses a configured nominal rehearsal floor (default $0.35$). Failure-profile weights
use smoothed failure, uniform mixing, and a maximum probability. If validation violates
$R_N\geq R_N^\star-0.10$, the guard:
\begin{enumerate}
\item raises nominal rehearsal toward its configured maximum;
\item reduces the failure-focus exponent;
\item logs a guard event and current constraint margin;
\item prevents an infeasible composite checkpoint from replacing the best feasible one.
\end{enumerate}
Reset sampling is split into a nominal group, a dedicated pass-required cooperation group
(default probability $0.20$), and a randomized robustness group. Only the last group draws a
failure-weighted perturbation profile.
The configurable composite score combines nominal, robustness, cooperation, and constraint
penalty. Separate \texttt{best\_nominal}, \texttt{best\_cooperation},
\texttt{best\_composite}, and \texttt{final} checkpoint labels avoid one opaque selection rule.

\section{Evaluation gates}
\textbf{Gate A---calibration.} Every required calibration check passes at the declared episode
count. \textbf{Gate B---nominal team play.} The recurrent seed-3 policy must meet the locked
2-v-2, 3-v-2, pass-required cooperation, duration, and meaningful-action thresholds.
\textbf{Gate C---constrained
robustness.} CC-FDR improves the robustness aggregate, remains within $0.10$ of the nominal
reference, and does not reduce cooperative success beyond its margin.

Development uses seed 3 only. Thresholds and checkpoint rules are then frozen. Final experiments
use seeds 4, 5, and 6, with seed-level mean, standard deviation, and the raw per-seed table.

\section{Phase 3 seed-3 development record}
The authoritative compact record is
\path{reports/phase3_development_summary.json}. Full Gate-A and Gate-B summaries are also retained
in executed notebook cells 12 and 22. The current Mac archive contains an older
\path{runs/phase3_calibration} smoke artifact with 20 episodes per cell; it is not the full Gate-A
artifact and must not replace the notebook's 100-episode evidence.

\begin{center}
\scriptsize
\begin{tabular}{lrrrr}
\toprule Stage & Steps & Updates & Last N & Last C \\
\midrule
A & 507,904 & 62 & 0.550 & 0.700 \\
B & 1,007,616 & 123 & 0.675 & 0.525 \\
C & 1,507,328 & 184 & 0.975 & 1.000 \\
D & 2,007,040 & 245 & 1.000 & 1.000 \\
\bottomrule
\end{tabular}
\normalsize
\end{center}
These are abstract validation values from a single development actor, not independent seed
estimates. The Stage-D final update had entropy $0.5173$, approximate KL $0.00106$, clipping
fraction $0.00763$, and explained variance $0.8566$. Logs contain no non-finite update or high-KL
stop. Learning-rate-floor messages are declared scheduler diagnostics; the Matplotlib/TorchScript
deprecation warnings do not explain Gate B. A zero success field in an individual terminal
rollout row in Stage B is also not a complete evaluation; the separate 100-episode Pymunk
pass-required evaluation was $0.37$.

The full calibration passed all nine checks at 100 episodes per required cell. Pymunk 2-v-2 open
success was random $0.01$, double-chase $0.03$, and role $0.26$; direct-shot double-chase success
in pass-required play was zero. The throughput diagnostic selected 128 environments at $337.45$
transitions/s. Values at 256 and 512 were essentially flat while wall time increased, showing a
physics/CPU plateau rather than a GPU failure.

The frozen Stage-D best-nominal checkpoint was evaluated for 100 episodes per scenario with seed
base 350000:
\begin{center}
\scriptsize
\begin{tabular}{lrrr}
\toprule Gate-B check & Obs. & Min. & Pass \\
\midrule
2-v-2 open success & 0.49 & 0.55 & fail \\
3-v-2 open success & 0.75 & 0.35 & pass \\
pass-required coop. & 0.61 & 0.30 & yes \\
successful duration & 17.2 s & 8.0 s & pass \\
meaningful actions & 6.19 & 2.0 & yes \\
\bottomrule
\end{tabular}
\normalsize
\end{center}
The CLI's exit code 2 and the notebook's former assertion were therefore the intended fail-closed
gate response, not an evaluator or checkpoint exception. The original notebook cell placed Drive
sync after the assertion, so the raw JSON remained in the Colab runtime and notebook output. The
repaired cell now saves and syncs first, displays the checks without a red assertion traceback,
and stores \texttt{GATE\_B\_PASSED}. The rerun artifact and all 300 rows are now synchronized at
\path{runs/20260723_094519_phase3_recurrent_nominal_mappo_seed3/eval/phase3_gate_b.json}.
The CC-FDR cell independently reads the JSON and refuses to train unless it passed.

\subsection{Critical diagnosis}
The result is promising cooperation, not nominal graduation. Stage B's Pymunk pass-required
success rose from $0.37$ to Stage D's $0.61$, and the learned actor substantially exceeds the
role baseline in open play ($0.49$ versus $0.26$). Yet the locked gate is conjunctive and the
abstract-to-Pymunk gap is large.

The 2-v-2 aggregate is brittle by defender style: lane-block $0.333$, predictive $0.053$, press
$1.000$, and zonal $0.483$. The raw rows distinguish three mechanisms:
\begin{itemize}
\item Lane-block failures ($n=18$) average 106.5 pass attempts, 105.8 receptions, 0.56 shots, zero
shots on target, controlled-reception reward $37.0$, and total return $35.82$. Lane successes
($n=9$) average 66.4 passes, 64.2 receptions, 5.33 shots, and return $37.34$. Pass attempts and
success have correlation $-0.623$ within this style. The transition bookkeeping is working, but
the task reward pays every legitimate back-and-forth reception; it therefore permits a failed
cycle to rival a goal-scoring return.
\item Predictive failures ($n=18$) average 11.61 shots, zero on target, 7.39 out-of-bounds events,
and return $-19.18$. This is a finishing/ball-control failure, not pass cycling.
\item Zonal failures ($n=15$) average 62.1 collision events, 3.6 passes, and 1.13 shots, versus
21.2 collisions, 17.2 passes, and 5.5 shots in successes. Audit confirms that the counter
increments once per macro-action decision with attacker overlap, so sustained contact may count
repeatedly. No collision reward coefficient changes in Stage R.
\end{itemize}

The six synchronized videos show genuine carrier/support separation, completed passes, and
scoring/restarts. All are exactly 20 seconds, however, so they terminate before a 50-second match
and contain empty final-metric fields. The recorder also overwrites
\path{video_manifest.json} on each invocation; the surviving manifest lists only the final three
3-v-2 clips. Section 23 sampled pass-required lane/zonal behavior, while Section 24 happened to
sample zonal 3-v-2 behavior. Consequently the videos establish nontrivial behavior but are not
the matched predictive/lane failure audit the gate needs.

\section{Immediate work contract: Stage R nominal mastery}
\textbf{Implementation status.} The contract below is implemented in
\path{configs/phase3_stage_r.yaml} and notebook Sections 25--38. Unit and tiny smoke evidence is
not a Stage-R performance result. Do not extend unchanged Stage D, change network
architecture, sweep PPO hyperparameters, train in Pymunk, add harder physics, or launch final
seeds. The GRU actor/critic has stable optimization and already solves abstract validation,
pass-required transfer, and 3-v-2 transfer. The immediate bottleneck is the football objective
and the small exposure to hard styles: under the Stage-D mixture, 2-v-2 predictive and lane-block
resets are only about 4\% and 6\% of all training resets.

\textbf{Step R0---repaired audit.} Evaluation, Gate B-R, and video now force a declared defender
style. Video manifests merge and deduplicate; partial clips retain a metric snapshot; terminal
mode and exact-seed replay are supported. Before learning, evaluate the frozen Stage-D policy for 100
episodes in every cell of four defender styles, two scenarios (2-v-2 open and pass-required), and
two backends. Use declared seeds and save raw rows. Record outcome, terminal reason, shots/on
target, out-of-bounds, collision semantics, pass efficiency, possession-chain progression, and
each reward component. Capture one predictive miss/OOB failure, one lane circulation failure, one
zonal collision failure, and matched successes.

\textbf{Step R1---locked reward invariant.} In Stage R, controlled-reception reward is zero;
retain it as a metric. Keep forward ball-progress, pass-advance, pass-to-goal, and terminal goal
signals. A backward pass or return to a previously reached chain threat/high-water mark must yield
zero new progress reward. Deterministic tests cover first progress, backward reception, A--B--A
and A--B--A--B circulation, incremental frontier, turnover/restart reset, pass-to-goal once, pass
without goal, and reward decomposition. The result is also saved as machine-readable gate input.
This is an immediate Stage-R schema change, not annealing and not a retroactive claim about A--D.

\textbf{Step R2---one focused training run.} Warm-start Stage-D \texttt{best\_nominal}, restoring
actor, critic, and normalizers but resetting optimizer, scheduler, rollout, and stage counters.
Train seed 3 for 500,000 abstract steps at learning rate $10^{-4}$ decaying to
$3\times10^{-5}$. Use scenario probabilities
\[
(0.50,\ 0.20,\ 0.15,\ 0.15)
\]
for 2-v-2 open, pass-required, 3-v-2 open, and 3-v-2 press; use defender probabilities
\[
(0.35,\ 0.35,\ 0.20,\ 0.10)
\]
for lane-block, predictive, zonal, and press. Validate on held-out, style-stratified abstract
seeds and select by unshaped worst-style/open competence subject to cooperation floors. Do not
select checkpoints on the Pymunk gate.

\textbf{Step R3---Gate B-R on untouched seeds.} Preserve the five original Gate-B criteria. Add
the predeclared development safeguards below; they supplement rather than rewrite the historical
gate:
\begin{itemize}
\item pass-required cooperation at least $0.55$ and 3-v-2 open success at least $0.70$;
\item 2-v-2 open worst defender style at least $0.25$, and mean lane/predictive success at least
$0.40$, using 100 fixed-style episodes per style;
\item for every style with both successful and failed rows, mean successful return exceeds mean
failed return by at least five; and
\item all no-progress reception-reward invariants pass in tests and the artifact records raw
reward components.
\end{itemize}
R0, validation, and Gate B-R use disjoint bases 360000, 365000, and 370000. If abstract Stage-R
safeguards pass but Pymunk
Gate B-R fails, the next single intervention is a calibrated abstract dynamics envelope; do not
train directly in Pymunk. If both abstract and Pymunk fail, rerun A--D plus R from scratch under
the corrected reward schedule. Only a Gate B-R pass authorizes CC-FDR and a replacement Section
25 input checkpoint.

\section{Throughput and batching}
Inference batches environment and attacker dimensions. The Phase 3 executor keeps physics
instances authoritative while performing tensor inference and rollout packing in batches. The
throughput diagnostic reports reset, environment-step, device-transfer, policy-inference,
actor/critic optimization, validation, checkpoint/logging I/O, effective transitions/s, and
memory for 4/8/16 environments by default, with capability
checks for 128--512 where hardware permits. Equivalence tests compare a single lane with the
ordinary environment under fixed states and actions. A faster result is not accepted if terminals,
rewards, masks, or observations diverge.

\section{Artifact and notebook discipline}
Runs use timestamped directories below \texttt{runs}, with experiment, algorithm, and seed in the
directory name. Training,
evaluation, plots, videos, resolved config, metadata, and logs are explicit subtrees. Latest
pointers prefer repository-relative \texttt{runs/<basename>} values and readers fall back to the
basename if an old Colab/Mac absolute path is stale. The workspace auditor checks metadata,
configuration, referenced checkpoints/evaluations, notebook run references, latest pointers,
duplicate run IDs, and incomplete artifacts without deleting anything.

The Phase 3 notebook has one initialization/install cell, one lightweight restore cell, and then
independently runnable audit, calibration, throughput, staged-training, evaluation, video, and
report cells. Training cells call the unbuffered CLI directly so live output remains visible; no
helper captures a subprocess until it finishes. Long loops use
\texttt{tqdm(..., file=sys.stdout)} and unbuffered notebook output.
Sections 25--38 form a self-contained Stage-R fast path after normal initialization/Drive restore:
they verify the authoritative decision JSON, restore the exact Stage-D model when absent, run R0,
save reward invariants, launch one visible Stage-R command, evaluate/Gate/video, update reports,
and sync. Section 39 independently requires a passing \texttt{phase3\_gate\_b\_r.json} and the
selected Stage-R checkpoint; historical Gate B cannot authorize it.

\section{Execution runbook}
\begin{enumerate}
\item Install editable package and run Ruff/pytest/workspace audit.
\item Run calibration smoke; inspect exact failed/passed reasons.
\item Run full calibration. Stop if Gate A fails.
\item Run 4/8/16 smoke throughput, then 32--512 lanes on the L4; select a stable measured count.
\item Train nominal seed 3 stage A; evaluate and resume B, C, then D only if each stage is healthy
(completed).
\item Evaluate frozen nominal best in abstract and Pymunk; Gate B failed and the post-sync audit
identified reward/style failures (completed).
\item Stage-R implementation and smoke verification (completed; not scientific evidence).
\item Run R0, reward invariants, one 500,000-step seed-3 Stage-R job, held-out abstract
validation, and Gate B-R on untouched seeds (current step).
\item Start CC-FDR only after Gate B-R passes.
\item Evaluate Gate C and inspect constraint-guard history.
\item Freeze config/protocol. Run seeds 4, 5, and 6.
\item Regenerate reports from JSON artifacts and compile both PDFs.
\end{enumerate}

\section{Exact observation and state ledger}
The observation dimension is 112 for every active attacker. The layout below is a schema contract,
not a suggestion; changing a field requires a schema-version increment and invalidates direct
checkpoint resume.
\begin{center}
\scriptsize
\begin{tabular}{p{0.12\columnwidth}p{0.10\columnwidth}p{0.60\columnwidth}}
\toprule Indices & Width & Content \\
\midrule
0--9 & 10 & self pose, velocity, heading, possession, slot/roster \\
10--17 & 8 & relative ball, ball velocity, touch/pass-chain state \\
18--21 & 4 & goal vector and scenario flags \\
22--41 & 20 & two deterministic teammate slots, 10 values each \\
42--59 & 18 & three deterministic defender slots, 6 values each \\
60--66 & 7 & previous-action one-hot \\
67--73 & 7 & executed-action one-hot \\
74--80 & 7 & legal-action mask \\
81--84 & 4 & clock, restart, cooperation, profile flags \\
85--111 & 27 & reserved zero padding for compatible extensions \\
\bottomrule
\end{tabular}
\normalsize
\end{center}
Each teammate slot begins with a presence bit and includes relative pose/velocity, possession, and
the last delivered four-value message. Each defender begins with presence and ends with a goalie
bit. A missing entity contributes exactly zero. Noise is applied to perceived positions, never to
presence or stable identity. The policy does not receive hidden profile parameters.

The global state dimension is 102. Attacker slots occupy 0--23 (eight values each), defender slots
24--44 (seven each), ball position/velocity 45--48, possession one-hot 49--52, and scenario/roster,
clock, and cooperation flags 53--58. Indices 59--101 are reserved. The centralized critic sees
world-frame values and masks, but its recurrent state and output are discarded during deployment.

\section{Action and transition ledger}
The seven categorical actions retain Phase 2 names: approach ball, dribble left, dribble right,
shoot, pass to teammate, move to support, and hold/face ball. Shoot and pass are masked unless the
agent currently possesses the ball. Pass is masked if no teammate is active or its predicted
receiver point leaves the field. Illegal requests from
a baseline or caller are counted and converted to hold; learned logits are masked before sampling,
so they should not generate such requests.

The exported TorchScript deployment wrapper accepts observation, recurrent state, and the current
legal-action mask, applies the same invalid-logit rule, and returns masked logits plus the next
hidden state. Deployment therefore cannot accidentally omit a mask used during training and
evaluation.

At each macro decision, a legal requested action enters the configured FIFO. The executed command
is repeated for the configured low-level integration count. Action latency of two therefore
executes hold, hold, then the first requested command. Reset fills queues with hold. A match restart
also clears the action queue, because a stale kick should not cross a restart, while recurrent
policy state remains intact because the multi-restart match episode did not terminate.

Abstract motion applies acceleration-bounded kinematics and explicit ball drag. Pymunk creates
separate circular rigid bodies and advances its own space with substeps. Both constrain players to
the playable rectangle, while the ball may cross goal or out-of-bounds lines. Swept defender-ball
checks prevent high-speed kicks tunnelling through a scripted defender. Defender possession
challenges and clearances are logged through last-touch state.

The pass-required scenario rejects a direct goal until a valid reception with configured forward
advance has occurred. A valid pass requires lane clearance, a distinct target, and reception inside
the window. A goal before the minimum-duration boundary is returned to play rather than counted;
this ensures the calibration cannot pass using an immediate kickoff exploit. Organized
lane-block/zonal and aggressive press/predictive styles create different post-pass openings without
changing the gate by method.

\section{Rollout tensor ledger}
For rollout length $T$, environment count $E$, and maximum attacker count $A=3$, the trainer saves:
\begin{center}
\small
\begin{tabular}{p{0.57\columnwidth}p{0.27\columnwidth}}
\toprule Tensor & Shape \\
\midrule observations & $T\times E\times A\times112$ \\
global states & $T\times E\times102$ \\
action masks & $T\times E\times A\times7$ \\
valid slots/actions/log-probabilities & $T\times E\times A$ \\
team rewards/values/returns & $T\times E$ \\
actor hidden snapshots & $T\times E\times A\times H$ \\
critic hidden snapshots & $T\times E\times H$ \\
continuation, terminal, truncation & $T\times E$ \\
\bottomrule
\end{tabular}
\normalsize
\end{center}
The hidden snapshot precedes the action at that time index. An episode reset zeros both actor and
critic hidden states and sets continuation to zero for the next observation. An in-match goal
restart changes world state but not the continuation mask. Inactive attacker rows have a valid
mask of zero and contribute neither policy loss nor entropy.

Actor chunks are keyed by environment and stable attacker slot. Critic chunks are keyed by
environment. A chunk starts from its recorded hidden snapshot; continuation masks reset recurrent
state if a later episode boundary lies inside the chunk. Short terminal chunks are padded and
masked. Advantages normalize only over active entries. Value loss and explained variance operate
at team/environment granularity. This avoids treating three padded attacker slots as three
independent team rewards.

\section{Checkpoint and stage invariants}
A Phase 3 checkpoint contains explicit checkpoint, observation, and state schema versions; actor
and critic weights; both optimizers; normalization moments; stage/update/step counters; best-score
records; last validation; curriculum; Python/NumPy/Torch RNG; and the resolved configuration.
Resume rejects a recurrent hidden-size or schema mismatch. Nominal Stage B resumes Stage A
optimizer/counters and trains only to its cumulative target; Stage C and D behave analogously.

Stage R uses warm start rather than ordinary A--D resume. It requires Stage-D provenance, copies
actor, critic, and both normalizers, and resets optimizer/scheduler/update/rollout state. It logs
the source checksum, reward schema, fixed mixes, seed blocks, floors, and reset declarations.
Its lexicographic best checkpoint first enforces cooperation \(0.50\) and 3-v-2 \(0.65\), then
maximizes the minimum of lane/predictive abstract success, four-style mean, overall success, and
cooperation. Pymunk never enters checkpoint selection.

CC-FDR similarly uses warm start, but now requires a passing Gate-B-R authorization artifact and
a Stage-R checkpoint. It reads the frozen nominal validation as the constraint reference; a
checkpoint without that reference is rejected. Starting
\texttt{phase3\_cc\_fdr.yaml} with neither \texttt{--warm-start} nor \texttt{--resume} is an error.
Historical A--D configurations still require the non-smoke calibration JSON. Stage R instead
requires the authoritative development summary to retain Gate A pass, historical Gate B fail,
and CC-FDR lock.

Learning-rate decay has actor and critic minima. Reaching a minimum logs a warning rather than
silently decaying useful updates to zero. PPO logs approximate KL, clipping fraction, entropy,
gradient norms, rollout/update time, and effective transitions per second. Non-finite loss or
gradient raises immediately and triggers an emergency checkpoint attempt.

\section{Calibration interpretation guide}
A smoke calibration contains 20 episodes per cell and must never authorize training even when all
observed checks pass. Its purpose is deterministic code coverage and detection of gross
permissiveness. The full gate uses at least 100 episodes in each of 12 required
method--simulator--scenario cells.

Random Pymunk success above 0.05 usually means that uncontrolled kick/dribble sequences can cross
the goal too easily. Double-chase above 0.20 means direct approach-and-shoot remains sufficient.
Role success below 0.20 means the pass/reception or scripted movement is broken or too difficult;
above 0.55 means the opponent/task is too predictable. A role--random gap below 0.15 means the
benchmark cannot distinguish structured coordination. Nonzero double-chase in pass-required play
means the scenario's direct-shot control is invalid. A short median or any cell above 0.85 is an
additional saturation warning.

Calibration code writes all episode rows, not only gate values. Before changing geometry, inspect
terminal reason, pass attempts/validity/receptions, defender style, time, action switches,
possession switches, redundant chase, and return. A change is acceptable only when it improves the
intended discrimination in both physics backends and passes the same locked thresholds.

\section{Development evaluation ledger}
Gate B evaluates frozen best-nominal weights in Pymunk on 2-v-2 open, 3-v-2 open, and 2-v-2
pass-required scenarios. The required values are respectively 0.55 success, 0.35 success, and 0.30
cooperative success, plus an eight-second successful median and at least two meaningfully used
actions. All are computed from episode rows. Abstract results remain useful learning diagnostics
but cannot substitute for this transfer-facing gate.

Gate B-R preserves all five historical Gate-B checks and adds Pymunk and abstract safeguards:
pass-required cooperation \(0.55\), 3-v-2 success \(0.70\), worst fixed-style 2-v-2 success
\(0.25\), lane/predictive mean \(0.40\), and a successful-minus-failed return gap of five whenever
both outcomes occur. All-success/all-failure style cells are explicitly not applicable for the
gap. The gate also requires the reward-invariant JSON. Its 14 cells use base 370000 and retain raw
rows; historical Gate B at 350000 is never overwritten.

Gate C evaluates recurrent nominal and CC-FDR on identical seeds. Profile mean averages the
declared non-nominal profiles after first reducing each profile. Grid AUC averages the declared
delay-low, delay-high, localization, and combined cells. Required contrasts are profile improvement
at least 0.10, nominal drop at most 0.10, grid drop at most 0.05, and cooperative drop at most zero.
The development seed remains one selected actor and is never included as an independent final
seed.

When final confirmation becomes authorized, seed 4, seed 5, and seed 6 are launched as separate
supervised jobs. Their evaluation ranges must be new, and tables first reduce to one metric per
trained seed. Report mean and sample standard deviation together with the raw three rows. Do not
bootstrap episodes and label the resulting interval as training-seed uncertainty.

\section{Artifact audit interpretation}
The workspace auditor is intentionally read-only. An error includes duplicate run identity or a
missing resolved configuration. Warnings include running/incomplete metadata, stale latest
pointers, missing referenced lightweight artifacts, and notebook references whose raw run
directory is absent locally. The last category is expected for the current Phase 2 seed-1/2
archive, but must remain visible rather than being silently called complete.

New latest pointers store a repository-relative \texttt{runs} path when the configured output root is the
repository's ordinary \texttt{runs}. The resolver first tries that portable value, then the pointer
location, then basename recovery. Consequently a historical
\path{/content/robot-soccer-transfer/runs/} pointer can resolve on a Mac by basename if the same
run has been synchronized. Drive pointers remain portable and metadata is copied last as the run's
completion marker.

Colab-to-Drive sync always includes checkpoints/models. Drive-to-Mac sync excludes checkpoint
directories and known heavy suffixes unless explicitly requested. This is asymmetric by design:
Colab needs resume state, while the Mac context needs configs, JSON/CSV/logs, plots, reports, and
videos. Sync merges and never deletes destination-only artifacts.

\section{Failure response matrix}
\begin{center}
\small
\begin{tabular}{p{0.27\columnwidth}p{0.57\columnwidth}}
\toprule Observation & Required response \\
\midrule
Gate A fails & stop; inspect episode rows; repair geometry/baseline; retain thresholds \\
non-finite recurrent update & stop; keep emergency checkpoint/log; reproduce in smoke \\
KL repeatedly high & reduce update aggressiveness from declared development config, then rerun seed 3 \\
entropy collapses before success & inspect masks/reward/action counts; do not add final seeds \\
passes selected but no receptions & audit target speed, lane validity, receiver behavior, and window \\
reception credit without transition & correctness bug; invalidate affected run and add regression test \\
Gate B any criterion fails & sync the full artifact; keep CC-FDR/harder platforms disabled;
diagnose frozen-policy events before retraining \\
Gate B-R abstract pass/Pymunk fail & stop; recommend only an evidence-targeted abstract dynamics
envelope \\
Gate B-R abstract and Pymunk fail & stop; recommend a future A--D plus R rerun under reward schema
2 \\
CC-FDR violates floor & verify guard events; raise rehearsal or stop/restore feasible checkpoint \\
Gate C profile gain absent & report negative mechanism result; do not cherry-pick a profile \\
artifact cannot regenerate table & remove table from claim until raw artifact is restored \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\section{Stop conditions}
Stop rather than spend the next budget if calibration fails, random/direct-shot succeeds too often,
role play is saturated, gradients are non-finite, PPO KL repeatedly exceeds the threshold, action
entropy collapses before competence appears, cooperative credit fires without a receiver
transition, nominal floor guard remains active for consecutive evaluations, or raw artifacts
cannot regenerate a reported table. Do not respond by weakening gates.

\section{Claim template after future execution}
If all gates pass, the permissible claim is narrow: recurrent parameter-shared MAPPO learned
repeatable team play in calibrated small-sided scenarios, and competence-constrained
failure-directed fine-tuning improved a locked cross-physics robustness metric without exceeding a
declared nominal-loss margin. If any gate fails, report the mechanism failure and seed variability;
do not substitute videos, training return, or an easier scenario for the locked outcome.

The current permissible Phase 3 claim is narrower: one development-seed recurrent actor learned
substantial, visually recognizable cooperative small-sided behavior and passed four of five
transfer-facing competence checks. It also exposed distinct style failures and a controlled-
reception reward that can favor circulation over scoring. Nominal-mastery Stage R is required;
its implementation is ready but its scientific run is absent. The actor has not advanced to
robust fine-tuning or confirmation.

\end{document}
